---
editor_options: 
  markdown: 
    wrap: 72
---

# Logic and Vectors {#log_vec}

In this chapter, we will introduce you to the basics of logic in a
computer. Logic is the branch of mathematics that deals with the
principles and methods of reasoning and inference. Logic is essential
for programming, as it allows us to create and manipulate expressions
and statements that can be evaluated as true or false. Logic also
enables us to control the flow and structure of our programs, and to
perform various operations and calculations on data.

## Comparisons

One of the most fundamental aspects of logic is comparison. Comparison
is the process of examining two or more objects or values and
determining their relationship or difference. Comparison can be used for
various purposes, such as checking the validity of an input, sorting a
list of items, or finding the optimal solution to a problem. In this
section, we will show you how to compare objects and values in both R
and Python, and how to use the results of comparison in your programs.

First, let's see how to check for equality in R and Python. Equality
means that two objects or values are exactly the same in every aspect.
To check for equality, we use the operator `==`, which returns `TRUE` if
the objects or values are equal, and `FALSE` if they are not. For
example, we can compare two strings, two integers, or two doubles, as
shown below:

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r}
# Comparing two strings
"Hello" == "Hello"

"Hello" == "hello"


# Comparing two integers
2 == 2

2 == 3


# Comparing two doubles
3.14 == 3.14

3.14 == 3.1415

```

</div>

<div>

**python**

```{python}
# Comparing two strings
"Hello" == "Hello"

"Hello" == "hello"


# Comparing two integers
2 == 2

2 == 3


# Comparing two doubles
3.14 == 3.14

3.14 == 3.1415

```

</div>
:::

As you can see, the syntax and the output are very similar in both
languages, except that R uses uppercase letters for `TRUE` and `FALSE`,
while Python uses lowercase letters for `True` and `False`. Note that
the comparison is case-sensitive, meaning that uppercase and lowercase
letters are considered different. Also note that the comparison is
precise, meaning that even a small difference in the decimal places can
make two doubles unequal.

Next, let's see how to check for inequality in R and Python. Inequality
means that two objects or values are not exactly the same in every
aspect. To check for inequality, we use the operator `!=`, which returns
`TRUE` if the objects or values are not equal, and `FALSE` if they are.
For example, we can compare two strings, two integers, or two doubles,
as shown below:

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r}
# Comparing two strings
"Hello" != "Hello"

"Hello" != "hello"


# Comparing two integers
2 != 2

2 != 3


# Comparing two doubles
3.14 != 3.14

3.14 != 3.1415

```

</div>

<div>

**python**

```{python}
# Comparing two strings
"Hello" != "Hello"

"Hello" != "hello"


# Comparing two integers
2 != 2

2 != 3


# Comparing two doubles
3.14 != 3.14

3.14 != 3.1415

```

</div>
:::

As you can see, the syntax and the output are very similar in both
languages, except that R uses uppercase letters for `TRUE` and `FALSE`,
while Python uses lowercase letters for `True` and `False`. Note that
the comparison is case-sensitive, meaning that uppercase and lowercase
letters are considered different. Also note that the comparison is
precise, meaning that even a small difference in the decimal places can
make two doubles unequal.

We can also compare the content of variables in both R and Python. A
variable is a name that refers to a value or an object that is stored in
the memory. To assign a value or an object to a variable, we use the
operator `<-` in R, and the operator `=` in Python. For example, we can
assign the string `"Hello"` to the variable `x` in both languages, as
shown below:

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r}
# Assigning a string to a variable
x <- "Hello"
```

</div>

<div>

**python**

```{python}
# Assigning a string to a variable
x = "Hello"
```

</div>
:::

To compare the content of variables, we use the same operators as
before, `==` for equality and `!=` for inequality. For example, we can
compare the variable `x` with another string, another variable, or
itself, as shown below:

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r}
# Comparing a variable with a string
x == "Hello"

x == "hello"


# Comparing a variable with another variable
y <- "World"
x == y

y <- "Hello"
x == y


# Comparing a variable with itself
x == x

```

</div>

<div>

**python**

```{python}
# Comparing a variable with a string
x == "Hello"

x == "hello"


# Comparing a variable with another variable
y = "World"
x == y

y = "Hello"
x == y


# Comparing a variable with itself
x == x

```

</div>
:::

As you can see, the syntax and the output are very similar in both
languages, except that R uses uppercase letters for `TRUE` and `FALSE`,
while Python uses lowercase letters for `True` and `False`. Note that
the comparison is case-sensitive, meaning that uppercase and lowercase
letters are considered different.

In the next section, we will introduce you to another important aspect
of logic: logical arithmetic. Logical arithmetic is the process of
combining and manipulating logical values, such as `TRUE` and `FALSE`,
using logical operators, such as `AND` and `OR`. Logical arithmetic can
be used for various purposes, such as creating complex conditions,
testing multiple hypotheses, or performing set operations. In this
section, we will show you how to use logical arithmetic in both R and
Python, and how to use the results of logical arithmetic in your
programs.

## Logical Arithmetic

Logical arithmetic is the process of combining and manipulating logical
values, such as `TRUE` and `FALSE`, using logical operators, such as
`AND` and `OR`. Logical arithmetic can be used for various purposes,
such as creating complex conditions, testing multiple hypotheses, or
performing set operations. In this section, we will show you how to use
logical arithmetic in both R and Python, and how to use the results of
logical arithmetic in your programs.

The two most common logical operators are `AND` and `OR`. The `AND`
operator returns `TRUE` if both operands are `TRUE`, and `FALSE`
otherwise. The `OR` operator returns `TRUE` if either or both operands
are `TRUE`, and `FALSE` otherwise. For example, we can use the `AND` and
`OR` operators to combine two logical values, such as `TRUE` and
`FALSE`, as shown below:

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r}
# Using the AND operator
TRUE & TRUE

TRUE & FALSE

FALSE & TRUE

FALSE & FALSE


# Using the OR operator
TRUE | TRUE

TRUE | FALSE

FALSE | TRUE

FALSE | FALSE

```

</div>

<div>

**python**

```{python}
# Using the AND operator
True and True

True and False

False and True

False and False


# Using the OR operator
True or True

True or False

False or True

False or False

```

</div>
:::

As you can see, the syntax and the output are very similar in both
languages, except that R uses `&` and `|` for `AND` and `OR`, while
Python uses `and` and `or`. Note that the order of the operands does not
matter, as `AND` and `OR` are commutative operators.

We can also use a table to show the possible outcomes of the `AND` and
`OR` operators, as shown below:

| a     | b     | a AND b | a OR b |
|-------|-------|---------|--------|
| TRUE  | TRUE  | TRUE    | TRUE   |
| TRUE  | FALSE | FALSE   | TRUE   |
| FALSE | TRUE  | FALSE   | TRUE   |
| FALSE | FALSE | FALSE   | FALSE  |

As you can see, the table has four rows, corresponding to the four
possible combinations of the operands a and b. Again, the table is easy to build, as it shows a AND b is only TRUE when both A and B are TRUE, a OR
b is only FALSE when both A and B are FALSE.

### Precendence of operators
In logical operations, as in general algebra, the precedence of operators determines the order in which operations are performed:

1. **Logical NOT**: This operator has the highest precedence and is performed first¹⁴.
2. **Logical AND**: This operator has the next highest precedence after Logical NOT¹⁵.
3. **Logical OR**: This operator has lower precedence than Logical AND¹⁵.

So, in an expression without parentheses, the NOT operation is performed first, followed by the AND operation, and finally the OR operation. 

For example, in the expression `NOT p AND q AND r`, the operations would be performed in the following order:

1. `NOT p` is performed first due to the highest precedence of `NOT`.
2. Then, `(NOTp) AND q` is performed due to the next highest precedence of `AND`.
3. Finally, `((NOT p) AND q) OR r` is performed due to the lowest precedence of `OR`.

Parentheses can be used to change this order. For example, in the expression `p ∧ (q ∨ r)`, the operation `q ∨ r` is performed first due to the parentheses, even though `∨` has lower precedence than `∧`.

For example, consider
the following expression:

`TRUE | TRUE & FALSE`

Without parentheses, in this expression the `AND` operator has higher
precedence than the `OR` operator, so it is evaluated first. The result
is then combined with the `OR` operator. The expression is equivalent
to:

`TRUE | (TRUE & FALSE)`

However, if we use parentheses to group the operands differently, we can
change the order of evaluation and the result of the expression. If we use parentheses to group the first and the last operands, we get:

`(TRUE | TRUE) & FALSE`

The value of this expression is in fact `FALSE`. Now, the `OR` operator is evaluated first, giving `TRUE` and the result is then combined with the `AND` operator, giving `FALSE`. 

#### Exercise
To test your understanding of logical arithmetic, try to find what is the value of:

-  `TRUE | TRUE & FALSE`?
-  `FALSE | FALSE & TRUE`?
-  `(TRUE & FALSE) | (FALSE & TRUE)`?
-  `(TRUE | TRUE) & FALSE`?
-  `(TRUE | FALSE) & (FALSE | TRUE)`?

You should use the console in R or Python to check your answers.

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r include=F}
# Checking the answers
TRUE | TRUE & FALSE
FALSE | FALSE & TRUE
(TRUE & FALSE) | (FALSE & TRUE)
(TRUE | TRUE) & FALSE
(TRUE | FALSE) & (FALSE | TRUE)
```

</div>

<div>

**python**

```{python include=F}
# Checking the answers
True and False or True
False or False and True
(True and False) or (False and True)
True and (False or True)
(True or False) and (False or True)
```

</div>
:::

Did you get the correct answers? If not, don't worry. You can review the
table and the examples above, and try to understand the logic behind
each expression. Remember that the `AND` operator has higher precedence
than the `OR` operator, which means that it is evaluated before the `OR`
operator. You can also use parentheses to group the operands and change
the order of evaluation.

## Flow

Another important aspect of logic is flow. Flow is the order and direction in which the instructions of a program are executed. Flow can be controlled by using conditional statements, such as if - else if - else. Conditional statements allow us to execute different blocks of code depending on the result of a logical expression or condition. For example, we can use conditional statements to perform different actions based on the value of a variable, the input of the user, or the outcome of a calculation.

The simplest form of a conditional statement is the if statement. The if statement checks a single condition, and executes a block of code if the condition is true. For example, we can use an if statement to print a message if a variable is positive, as shown below:

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r}
# Assigning a value to a variable
x <- 10

# Using an if statement
if (x > 0) {
  print("x is positive")
}
# x is positive
```

</div>

<div>

**python**

```{python}
# Assigning a value to a variable
x = 10

# Using an if statement
if x > 0:
  print("x is positive")
# x is positive
```

</div>
:::

As you can see, the syntax and the output are very similar in both languages, except that Python uses a colon and indentation to indicate the start and the end of the block of code. Note that the if statement only executes the block of code if the condition is true, and does nothing otherwise.

Sometimes, we may want to check more than one condition, and execute different blocks of code for each condition. For this, we can use the if - else if - else statement. The if - else if - else statement checks multiple conditions in order, and executes the first block of code that corresponds to a true condition. If none of the conditions are true, the else block of code is executed. For example, we can use an if - else if - else statement to print a message based on the sign of a variable, as shown below:

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r}
# Assigning a value to a variable
x <- 10

# Using an if - else if - else statement
if (x < 0) {
  print("x is negative")
} else if (x == 0) {
  print("x is zero")
} else {
  print("x is positive")
}
# x is positive
```

</div>

<div>

**python**

```{python}
# Assigning a value to a variable
x = 10

# Using an if - else if - else statement
if x < 0:
  print("x is negative")
elif x == 0:
  print("x is zero")
else:
  print("x is positive")
# x is positive
```

</div>
:::

As you can see, the syntax and the output are very similar in both languages, except that R uses `else if` and Python uses `elif` for the second condition. Note that the order of the conditions is important, as only the first condition that is true is executed. If none of the conditions are true, the else block is executed.

Sometimes, we may want to check two or more independent conditions, and execute different blocks of code for each condition. For this, we can use two or more if statements one after another. This is different from using an if - else if - else statement, as each if statement is evaluated separately, and more than one block of code can be executed. For example, we can use two if statements to print a message based on the divisibility of a variable by 2 and 3, as shown below:

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r}
# Assigning a value to a variable
n <- 12

# Using two if statements
if (n %% 2 == 0) {
  print("n is divisible by 2")
}
if (n %% 3 == 0) {
  print("n is divisible by 3")
}
# n is divisible by 2
# n is divisible by 3
```

</div>

<div>

**python**

```{python}
# Assigning a value to a variable
n = 12

# Using two if statements
if n % 2 == 0:
  print("n is divisible by 2")
if n % 3 == 0:
  print("n is divisible by 3")
# n is divisible by 2
# n is divisible by 3
```

</div>
:::

As you can see, the syntax and the output are very similar in both languages, except that R uses `%%` for the modulo operator, while Python uses `%`. Note that both if statements are executed, as both conditions are true. If we used an if - else if - else statement instead, only the first condition would be executed, and the second condition would be ignored.

Sometimes, we may want to check a condition within another condition, and execute different blocks of code for each combination. For this, we can use a nested if statement. A nested if statement is an if statement inside another if statement. A nested if statement can have multiple levels of nesting, and can also include else if and else blocks. For example, we can use a nested if statement to print a message based on the value and the sign of a variable, as shown below:

::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}
<div>

**R**

```{r}
# Assigning a value to a variable
x <- -10

# Using a nested if statement
if (x > 0) {
  if (x > 10) {
    print("x is positive and greater than 10")
  } else {
    print("x is positive and less than or equal to 10")
  }
} else {
  if (x < -10) {
    print("x is negative and less than -10")
  } else {
    print("x is negative and greater than or equal to -10")
  }
}
# x is negative and greater than or equal to -10
```

</div>

<div>

**python**

```{python}
# Assigning a value to a variable
x = -10

# Using a nested if statement
if x > 0:
  if x > 10:
    print("x is positive and greater than 10")
  else:
    print("x is positive and less than or equal to 10")
else:
  if x < -10:
    print("x is negative and less than -10")
  else:
    print("x is negative and greater than or equal to -10")
# x is negative and greater than or equal to -10
```

</div>
:::

As you can see, the syntax and the output are very similar in both languages, except that Python uses a colon and indentation to indicate the start and the end of each block of code. Note that the nested if statement checks the outer condition first, and then checks the inner condition if the outer condition is true. Note also that the nested if statement can have multiple levels of nesting, and can also include else if and else blocks.
